An efficient composition of bidirectional programs by tupling and lazy updates

### 2020-02-19 paper/1.introduction.tex
add todo list

\begin{figure}[!t]
  \begin{minipage}{0.3\textwidth}
    \centering
    \includegraphics[height=3.5cm]{./fig/fig1.eps}
    \caption{Evaluating $phead$}
    \label{fig:eval-phead}
  \end{minipage}\hfill
  \begin{minipage}{0.7\textwidth}
    \centering
    \includegraphics[height=3.5cm]{./fig/fig2.eps}
    \caption{Evaluating $phead \circ phead \circ phead$}
    \label{fig:eval-comp-phead}
  \end{minipage}
\end{figure}
\vspace*{-5.5mm}

@@

\minusvspace
\begin{align}
get_{bx_1 \circ bx_2}~s &= get_{bx_2} (get_{bx_1}~s)\\
put_{bx_1 \circ bx_2}~s~v' &= put_{bx_1} s~(put_{bx_2}~ (get_{bx_1}~s)~v')
\end{align}
\minusvspace

@@

To treat large inputs, we propose the second method based on memoization: ``keeping \emph{complements in a closure} and using them when needed''. The key idea is straightforward: Complements are smaller
% program fragments
than
% the original
intermediate states.
%Therefore this strategy solves the previous two problems.
%Readers might already know that it is a hard problem to obtain complements in general. In our work, thanks to the following finding, it is not hard:
%
%\vspace{2mm}
%In very-well behaved \cite{Foster:2007:CBT:1232420.1232424} BX,% programs,
%$put$ is a complement function for $get$.
%, $get$ can be a complement function for $put$.
%\vspace{2mm}
%
For obtaining complements, we tuple $put$ and $get$, and produce a new function $pg$. Because $put$ produces new complements for $get$, we can shrink the size.
Let us reconsider our example $phead \circ phead \circ phead$ in Figure \ref{fig:eval-comp-phead-2}, where $c_1$ and $c_2$ are complements and $d_1$ and $d_2$ are valid views for $s$ and~$i_1$. Here, two points are worth noting. First, after evaluating the leftmost $pg$, the original source $s$ need not be kept because its complete contents are in $c_1$ and~$i_1$. Second, the complements are smaller than the intermediate states in Figure~\ref{fig:eval-comp-phead}.
%This evaluation looks better than previous two strategies: this does not require repeated evaluation and require smaller storage than the original sources.
Actually, this simple
% combined
$pg$ alone is not yet effective for left associative compositions because it requires two more $put$s, which can be seen on the right side of Figure \ref{fig:eval-comp-phead-2}. To achieve an efficient evaluation, we combine two techniques, lazy update and lazy evaluation. We explain this second method and all optimizations in Section~\ref{sect:xpg}.

>>>>

\begin{figure}[!t]
  \begin{minipage}{0.3\textwidth}
    \centering
    \includegraphics[height=3.5cm]{./fig/fig1.eps}
    \caption{Evaluating $phead$}
    \label{fig:eval-phead}
  \end{minipage}\hfill
  \begin{minipage}{0.7\textwidth}
    \centering
    \includegraphics[height=3.5cm]{./fig/fig2.eps}
    \caption{Evaluating $phead \circ phead \circ phead$}
    \label{fig:eval-comp-phead}
  \end{minipage}
\end{figure}
%\vspace*{-5.5mm}

@@

\minusvspace
\begin{align}
get_{bx_1 \circ bx_2}~s &= get_{bx_2} (get_{bx_1}~s) \AlignFootnote{Note that the order of composition is the reverse order compared to function composition. We use this order, because it is standard in BX community.}\\
put_{bx_1 \circ bx_2}~s~v' &= put_{bx_1} s~(put_{bx_2}~ (get_{bx_1}~s)~v')
\end{align}
\minusvspace

@@

\textcolor{red}{TODO-B3,C1: add explanation of complements.}
To treat large inputs, we propose the second method based on memoization: ``keeping \emph{complements in a closure} and using them when needed''. The key idea is straightforward: Complements are smaller
% program fragments
than
% the original
intermediate states.
%Therefore this strategy solves the previous two problems.
%Readers might already know that it is a hard problem to obtain complements in general. In our work, thanks to the following finding, it is not hard:
%
%\vspace{2mm}
%In very-well behaved \cite{Foster:2007:CBT:1232420.1232424} BX,% programs,
%$put$ is a complement function for $get$.
%, $get$ can be a complement function for $put$.
%\vspace{2mm}
%
For obtaining complements, we tuple $put$ and $get$, and produce a new function $pg$. Because $put$ produces new complements for $get$, we can shrink the size.
Let us reconsider our example $phead \circ phead \circ phead$ in Figure \ref{fig:eval-comp-phead-2}, where $c_1$ and $c_2$ are complements and $d_1$ and $d_2$ are valid views for $s$ and~$i_1$. Here, two points are worth noting. First, after evaluating the leftmost $pg$, the original source $s$ need not be kept because its complete contents are in $c_1$ and~$i_1$. Second, the complements are smaller than the intermediate states in Figure~\ref{fig:eval-comp-phead}.
%This evaluation looks better than previous two strategies: this does not require repeated evaluation and require smaller storage than the original sources.
Actually, this simple
% combined
$pg$ alone is not yet effective for left associative compositions because it requires two more $put$s, which can be seen on the right side of Figure \ref{fig:eval-comp-phead-2}. To achieve an efficient evaluation, we combine two techniques, lazy update and lazy evaluation. We explain this second method and all optimizations in Section~\ref{sect:xpg}.

### 2020-02-19 paper/4.xpg.tex
tiny
small updates
add todo list

The first equality is simply based on the definitions of $pg$, $\putbxinline{Skip \ h}$ and $\getbxinline{Skip \ h}$. The second one tuples two results of $put$ and $get$ in the body of the if-expression. This is a trick since in some cases, the result of $pg$ may be undefined although the result is not undefined when evaluating $\getbxinline{Skip \ h}$. The last equality is relatively obvious.

@@

A restriction for $\pginline{Case}$ needs to be introduced here. We know that there is one entering condition and one exit condition when evaluating $\putbxinline{Case}$ as well as $\getbxinline{Case}$. If a tupling occurs, there will be 4 combinations from these conditions. This means two entering conditions of $\putbxinline{Case}$ and $\getbxinline{Case}$ are not always simultaneously satisfied. The evaluated branches are distinct in the $put$ and $get$ directions for combinations $((cond_{sv} \ s \ v) \ \&\& \ (not (cond_{s} \ s)))$ and $((not (cond_{sv} \ s \ v)) \ \&\& \ (cond_{s} \ s))$, which are restricted in this paper. This does not happen for the others which is used in the construction of $\pginline{Case}$.

@@

\smallvspace
    $\tab (ks, kv, s, v) \Leftarrow \cpg{bx}{\lambda \_.s_0}{id}{s_0}{v_0};\\
        \tab \qtab (s; v)$
\smallvspace

@@

\smallvspace
    $\tab (ks, kv, lks, lkv, s, v) \Leftarrow \kpg{bx}{\lambda \_.s_0}{id}{id}{id}{s_0}{v_0};\\
        \tab \qtab (lks \ s; lkv \ v)$
\smallvspace

@@

\textcolor{red}{
In $kpg$, basically, functions for the updates will be kept in $lks$ and $lkv$.
$\kpginline{Skip}$ and $\kpginline{Case}$ are special cases, because they require the actual values for evaluation. Therefore we evaluate the values by application of $lks \ s$ and $lkv \ v$ and update $lks$ and $lkv$ with the identity functions.
%Except $\kpginline{Skip}$ and $\kpginline{Case}$ where $s$ and $v$ hold actual values, functions for the computation will be kept in $ks'$ and $kv'$.
When evaluating $\kpginline{bx_1 \circ bx_2}$, $\underline{lks_1}$ and $\underline{s_1}$ in the result of the first assignment are still redundant but not evaluated.
\\
Additionally we did two optimizations in $kpg$. The first is in $\kpginline{bx_1 \times bx_2}$. The evaluation of $lks$ and $lkv$ will be done independently in two assignments using $\kpginline{bx_1}$ and $\kpginline{bx_2}$. There may be same computations in $fst \circ lks$ and $snd \circ lks$ as well as $fst \circ lkv$ and $snd \circ lkv$.
To remove duplicate evaluations, we evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$.
%It is possible to evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$ to remove the redundancy like that. One more thing,
Second is in $\kpginline{Case}$. We need to evaluate $lks \ s'$ and $lkv \ v'$ to check the \underline{fi} condition before returning the 6-tuple. Such evaluations can be done lazily to make programs run faster. We use the above small optimizations in our implementation.
}

\subsection{Combination of pg and kpg: xpg}
The most important thing when using $kpg$ to evaluate a composition $bx_1 \circ bx_2$ is keeping the dropped parts from the source in a function that are used to produce new source instead of using more recursive call. This is reflected in the first call $\kpginline{bx_1}$ and the last application in $\kpginline{bx_1 \circ bx_2}$. The second call $\kpginline{bx_2}$ just play a syntactic role while it requires many parameters. We can completely reduce this number to 2, including one for the source and one for the view as in a $pg$ call. This idea is realized by function $xpg$ as follows:

>>>>

The first equality is simply based on the definitions of $pg$, $\putbxinline{Skip \ h}$ and $\getbxinline{Skip \ h}$. The second one tuples two results of $put$ and $get$ in the body of the if-expression. This is a trick since in some cases, the result of $pg$ may be undefined although the result is not undefined when evaluating $\getbxinline{Skip \ h}$. The last equality is a function application.

@@

A restriction for $\pginline{Case}$ needs to be introduced here. We know that there is one entering condition and one exit condition when evaluating $\putbxinline{Case}$ as well as $\getbxinline{Case}$. If a tupling occurs, there will be 4 combinations from these conditions. This means two entering conditions of $\putbxinline{Case}$ and $\getbxinline{Case}$ are not always simultaneously satisfied. The evaluated branches are distinct in the $put$ and $get$ directions for combinations $((cond_{sv} \ s \ v) \ \&\& \ (not (cond_{s} \ s)))$ and $((not (cond_{sv} \ s \ v)) \ \&\& \ (cond_{s} \ s))$, which are restricted in this paper\footnote{\textcolor{red}{If the BX programs have efficient inverse way of evaluation, they are not affected by this restriction. Of course, some programs does not have such efficient way. An example restricted by this is evaluating $snoc$ with different length of a source and a view.}}. This does not happen for the others which is used in the construction of $\pginline{Case}$.

@@

\smallvspace
    $\tab (ks, kv, s, v) \Leftarrow \cpg{bx}{\lambda \_.s_0}{id}{s_0}{v_0};\\
        \tab \qtab (s, v)$
\smallvspace

@@

\smallvspace
    $\tab (ks, kv, lks, lkv, s, v) \Leftarrow \kpg{bx}{\lambda \_.s_0}{id}{id}{id}{s_0}{v_0};\\
        \tab \qtab (lks \ s, lkv \ v)$
\smallvspace

@@

\textcolor{red}{TODO-C6: RearrS, RearrV just keeping, Really evaluated is skip, product}
In $kpg$, basically, functions for the updates will be kept in $lks$ and $lkv$.
$\kpginline{Skip}$ and $\kpginline{Case}$ are special cases, because they require the actual values for evaluation. Therefore we evaluate the values by application of $lks \ s$ and $lkv \ v$ and update $lks$ and $lkv$ with the identity functions.
%Except $\kpginline{Skip}$ and $\kpginline{Case}$ where $s$ and $v$ hold actual values, functions for the computation will be kept in $ks'$ and $kv'$.
When evaluating $\kpginline{bx_1 \circ bx_2}$, $\underline{lks_1}$ and $\underline{s_1}$ in the result of the first assignment are still redundant but not evaluated.
\\
Additionally we did two optimizations in $kpg$. The first is in $\kpginline{bx_1 \times bx_2}$. The evaluation of $lks$ and $lkv$ will be done independently in two assignments using $\kpginline{bx_1}$ and $\kpginline{bx_2}$. There may be same computations in $fst \circ lks$ and $snd \circ lks$ as well as $fst \circ lkv$ and $snd \circ lkv$.
To remove duplicate evaluations, we evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$.
%It is possible to evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$ to remove the redundancy like that. One more thing,
Second is in $\kpginline{Case}$. We need to evaluate $lks \ s'$ and $lkv \ v'$ to check the \underline{fi} condition before returning the 6-tuple. Such evaluations can be done lazily to make programs run faster. We use the above small optimizations in our implementation.

\subsection{Combination of pg and kpg: xpg}
\textcolor{red}{TODO-D3: change explanation clearer.}
The most important thing when using $kpg$ to evaluate a composition $bx_1 \circ bx_2$ is keeping the dropped parts from the source in a function that are used to produce new source instead of using more recursive call. This is reflected in the first call $\kpginline{bx_1}$ and the last application in $\kpginline{bx_1 \circ bx_2}$. The second call $\kpginline{bx_2}$ just play a syntactic role while it requires many parameters. We can completely reduce this number to 2, including one for the source and one for the view as in a $pg$ call. This idea is realized by function $xpg$ as follows:

### 2020-02-19 paper/5.experiments.tex
add todo list

\textit{breverse} is defined in terms of \textit{bfoldr}. In the \textit{put} direction, it produces a reverse of the view list if its length is not larger than the length of the source list and, in the \textit{get} direction, produces a reverse of the source list. Note that compositions are by the recursions of \textit{breverse} and 
the number of compositions are dynamically determined by the length of the source list.

@@

Figure \ref{fig:result} 
% illustrates 
shows the evaluation times for each of the seven test cases using the three methods: $put$ in minBiGUL, $put_m$ in minBiGUL$_m$ and $xpg$.
We also did similar experiments with $pg$, $cpg$ and $kpg$, but their results are slower than the corresponding ones of $xpg$. The slowness was caused by the exponential number of $pg$ calls in the case of $pg$, redundant evaluations ($cpg$), and redundant overhead for constructing closures ($kpg$).
Then we simply omit these results in Figure \ref{fig:result}.
As we all know, $put$ works poorly for left associative compositions because of the number of reevaluated $get$s.
%The figure shows $put_m$ and $xpg$ are better than $put$ in some cases.
The left part of the figure contains tests using not-large inputs, and we see that $put_m$ is the fastest method for them. However if the input size is large enough, in the cases of the right part of the figure, $put_m$ will be slower quickly due to time for manipulating data in the table. At that time, $xpg$ is 
% proved to be
the most effective method.

>>>>

\textit{breverse} is defined in terms of \textit{bfoldr}, that appeared in Section \ref{sect:minbigul}. In the \textit{put} direction, it produces a reverse of the view list if its length is not larger than the length of the source list and, in the \textit{get} direction, produces a reverse of the source list. Note that compositions are by the recursions of \textit{breverse} and 
the number of compositions are dynamically determined by the length of the source list.

@@

\textcolor{red}{TODO-D4: emphasize overhead of memoization. it is not for free.}
Figure \ref{fig:result} 
% illustrates 
shows the evaluation times for each of the seven test cases using the three methods: $put$ in minBiGUL, $put_m$ in minBiGUL$_m$ and $xpg$.
We also did similar experiments with $pg$, $cpg$ and $kpg$, but their results are slower than the corresponding ones of $xpg$. The slowness was caused by the exponential number of $pg$ calls in the case of $pg$, redundant evaluations ($cpg$), and redundant overhead for constructing closures ($kpg$).
Therefore we simply omit these results in Figure \ref{fig:result}.
As we all know, $put$ works poorly for left associative compositions because of the number of reevaluated $get$s.
%The figure shows $put_m$ and $xpg$ are better than $put$ in some cases.
The left part of the figure contains tests using not-large inputs, and we see that $put_m$ is the fastest method for them. However if the input size is large enough, in the cases of the right part of the figure, $put_m$ will be slower quickly due to time for manipulating data in the table. At that time, $xpg$ is 
% proved to be
the most effective method.
\textcolor{red}{TODO-B2: add short evaluation result of memory.}

### 2020-02-19 paper/main.tex
Tiny
Capitalized \title, \titlerunning.
reduce some vspaces and add some vspaces
tiny changes in formulas: keywords --> normal text, redundant vars --> underline
small updates
rm \newcommand{getmh}
add todo list

\newcommand{\sif}[3]{\text{if } #1 \text{ then } #2 \text{ else } #3}
\newcommand{\product}[2]{#1 \times #2}
\newcommand{\tuple}[2]{(#1 :: #2)}
\newcommand{\rearrs}[3]{RearrS \ #1 \ #2 \ #3}
\newcommand{\rearrv}[3]{RearrV \ #1 \ #2 \ #3}
\newcommand{\casebx}[4]{Case \ #1 \ #2 \ #3 \ #4}

@@

\newcommand{\getmh}[2]{get_{mh} \, [\![#1]\!] \ #2}
\newcommand{\getmhinline}[1]{get_{mh} \, [\![#1]\!]}

@@

\newcommand{\xpg}[3]{xpg \, [\![#1]\!] (#2, #3)}
\newcommand{\xpginline}[1]{xpg \, [\![#1]\!]}

@@

\begin{document}
%
\title{
  % Strategies for efficient composition of bidirectional programs
  % by memoization and lazy updates
  %  tupling
  %  accumulating updates = lazy updates
  %  based on the idea of reversible computing  
  % \thanks{Supported by organization x.}  
  An Efficient Composition of Bidirectional Programs
  by Memoization and Lazy Update
}
%
\titlerunning{Strategies for efficient evaluation of BX programs}
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%

%\author{{Kanae Tsushima\inst{1}\orcidID{0000-0002-3383-3389}} \and
%  {Bach-Trong Nguyen\inst{1}\orcidID{0000-0002-6342-6020}} \and
%  {Robert Gl\"{u}ck\inst{2}\orcidID{0000-0001-6990-3935}} \and
%  {Zhenjiang Hu\inst{3}\orcidID{0000-0002-9034-205X}}}

\author{anonymous authors}

%\authorrunning{K. Tsushima et al.}

@@

%\institute{National Institute of Informatics, Japan
%  \email{\{k\_tsushima,bach\}@nii.ac.jp} \and
%  University of Copenhagen, Denmark
%  \email{glueck@di.ku.dk } \and
%  Peking University, China
%  \email{huzj@pku.edu.cn}}

>>>>

\newcommand{\sif}[3]{\textnormal{if} \ #1 \ \textnormal{then} \ #2 \ \textnormal{else} \ #3}
\newcommand{\product}[2]{#1 \times #2}
\newcommand{\tuple}[2]{(#1 :: #2)}
\newcommand{\rearrs}[3]{RearrS \ #1 \ #2 \ #3}
\newcommand{\rearrv}[3]{RearrV \ #1 \ #2 \ #3}
\newcommand{\casebx}[4]{Case \ #1 \ #2 \ #3 \ #4}

@@

@@

\newcommand{\xpg}[3]{xpg \, [\![#1]\!] (#2, #3)}
\newcommand{\xpginline}[1]{xpg \, [\![#1]\!]}
\newcommand{\smallvspace}{\vspace{0.3mm}}
\newcommand{\minusvspace}{\vspace*{-5.5mm}}
\newcommand{\minusvspacetwo}{\vspace*{-3.5mm}}

\newcommand{\match}[1]{\textnormal{match } #1 \ \textnormal{ with }}

\makeatletter
\newcommand{\AlignFootnote}[1]{%
    \ifmeasuring@
    \else
        \footnote{#1}%
    \fi
}
\makeatother

@@

\begin{document}
%
\title{
  % Strategies for efficient composition of bidirectional programs
  % by memoization and lazy updates
  %  tupling
  %  accumulating updates = lazy updates
  %  based on the idea of reversible computing  
  % \thanks{Supported by organization x.}  
  An Efficient Composition of Bidirectional Programs
  by Memoization and Lazy Update
}
%
\titlerunning{An Efficient Composition of BX Programs}
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%

\author{{Kanae Tsushima\inst{1}\orcidID{0000-0002-3383-3389}} \and \break
  {Bach-Trong Nguyen\inst{1}\orcidID{0000-0002-6342-6020}} \and \break
  {Robert Gl\"{u}ck\inst{2}\orcidID{0000-0001-6990-3935}} \and
  {Zhenjiang Hu\inst{3}\orcidID{0000-0002-9034-205X}}}

%\author{anonymous authors}

\authorrunning{K. Tsushima et al.}

@@

\institute{National Institute of Informatics, Japan
  \email{\{k\_tsushima,bach\}@nii.ac.jp} \and
  University of Copenhagen, Denmark
  \email{glueck@di.ku.dk } \and
  Peking University, China
  \email{huzj@pku.edu.cn}}
